---
title: "Data Summaries & Visualization"
author: "John Karuitha"
date: "3/18/2021"
output:
  beamer_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Describing and Exploring Categorical Data
- As noted, categorical or qualitative data shows categories/ buckets/ classes. 

- Examples; are height- Short, Medium, Tall; Age- 0 to 10 years, 11-15 years, 16-20 years, etc. 

- Note that the examples above could be quantitative e.g age of three students- 10, 12, 16 years. 

- You must distinguish qualitative and quantitative variables by looking at the context and presentation. 

## How can we summarise categorical variables
- We mostly use frequency tables or contigency tables

- Lets have an example as follows for the ages of senior high school students 

```{r}

(age <- tibble(name = c("Paul", "Catherine", "Oloo", "Etyang", "Nyaga", "Mwajuma"), 
              
              age = c("10-15", "16-20", "16-20", "16-20", "10-15", "21-25")))
```

## Summarising categorical variables
- Note that this data is categorical- students are presented in age groups and not real age. 

- Let us summarise the age groups in tabular format. 

```{r}
table(age$age) %>% knitr::kable(col.names = c("Age", "Freq"))
```
- We can also do relative frequencies as percentage of the occurrence of each age group. 
```{r}
prop.table(table(age$age))
```

## Contigency tables

- We typically summarise data using contingency tables. 

- Let us have a a look at another example.

```{r}
(age2 <- tibble(name = c("Paul", "Catherine", "Oloo", "Etyang", "Nyaga", "Mwajuma"), 
              
              age = c("10-15", "16-20", "16-20", "16-20", "10-15", "21-25"),
              
              sex = c("Male", "Female", "Male", "Male", "Male", "Female")))

```

```{r}
table(age2$sex)
```

- we can have a two way contingency table as follows.

```{r}
table(age2$age, age2$sex)
```

- Next we shall look at skewness and kurtosis.
- Measures of spread- sd, variance, IQR, ....

## Exercises

- You are provided with the following categorical dataset for ages of ten random people.

- Also remember how to write a character vector in R; Using quotes. 

```{r, echo = TRUE}
ages <- c("Toddler", "Teen", "Teen", "Adult", 
          
          "Adult", "Elder", "Adult", "Toddler", 
          
          "Teen", "Teen")

ages
```

- Draw a frequency table
- Draw a relative frequency table

#############################################
## Setting up a project/working directory in R


### Intro to R-Markdown
